Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging

### 2012-05-23 HRUIonGPU/jrnl_article.tex

\author{J. P. \AA{}sen, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{M. Shell is with the Department
of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}

@@

\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.

Capon beamforming has recently shown promising results in several studies. Although it improves the lateral resolution, widespread use has been limited by its computational complexity. Luckily, the calculation of the Capon weights is well suited for a GPU (Graphics Processing Unit) implementation, since the calculation is both independent across pixels, and partially intra-pixel.

In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing. Both types of smoothing are important for maintaining delay-and-sum-like speckle and at the same time keep the high resolution property. This is achieved using complex base band data, which reduces the size of each range line to a minimum. In addition, complex data results in complex weights; hence this is a correct implementation of Capon beamforming with all degrees of freedom preserved.

Delay and sum beamformer (DAS):
\begin{align}
z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where $M$ is the number of elements.

\subsection{Capon Beamforming}

Capon weights are found by solving the following minimization problem:
\begin{align}
&\min_{\vec{w}} E\{|z[n]|^2\} = \vec{w}^H\mat{R}\vec{w}\\
&\text{subjected to } \vec{w}^H\vec{1} = 1,
\end{align}
where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$.

The solution to this minimization problem is
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
\end{align}
where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed to broadside,  and $\vec{w} \in \mathbb{C}^M$.

Estimation of $\mat{R}[n]$: $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$.

In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ subarrays and $N_{avg}$ time samples.
\begin{align}
\mat{\hat{R}} = \frac{1}{(2N_{avg} + 1)K}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,
\end{align}
where $K = M-L+1$, $Y_{avg}\sim \tau/T_s$ ($\tau$ is the imaging pulse length and $T_s$ is the sampling period.) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.

Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, $\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}$. A weighting proportional to the output power is beneficial, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$, has previously shown good results in the literature (cite).

Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change.

\subsection{GPU Compute Model}
In this section we briefly discuss the principle behind GPU computing. CUDA by Nvidia has been our choice of GPU developer environment. However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper will apply for other massive parallel implementation environments like e.g. OpenCL. The CUDA compute model is based on execution of a kernel across a grid of compute threads. Each position in the grid holds a block of threads. Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. From a CUDA perspective the GPU consist of one or more stream multi processors (SM), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions. The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks. The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48. Hence, with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once. In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1. Up to a curtain point it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory. The SM has for CC 2.x 32K 32-bit long registers. When implementing algorithms on the GPU it is important to keep data as close to the core as possible. Utilizing available registers and shared memory to its maximum is therefore important.

@@

\section{Parallel Capon}\label{sec:meth}
\subsection{Building Multiple Sample Covariance Matrices}
Building $\hat{\mat{R}}$ using a sliding window approach across $K$ and $N_{avg}$. We have a limited amount of fast, near-core memory on the GPU. On NVIDIA architecture this memory is known as shared memory, and is restricted to 48KB per compute block. The maximum block size of 32x32, further restrict how large arrays we can handle inside a single compute block. It is therefore natural to restrict $L$ to a maxima of $32$ elements. Since $L \le M/2$, $M \le 64$. We can not afford to hold the full $\hat{\mat{R}}$ ($M = L$) when $\hat{\mat{R}} > 32$, because of limited amount of memory. ... Explain in detail how \texttt{buildR} is implemented in a kernel. ... Give more details on how much shared memory we have, and how it can be divided between compute blocks occupying one stream multiprocessor (SM).

>>>>

\author{J. P. \AA{}sen, J. I. Buskenes, C.-I. C. Nilsen, A. Austeng and S. Holm%\IEEEmembership{Member,~IEEE,}% <-this % stops a space
%\thanks{M. Shell is with the Department
%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
}

@@

%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.

\IEEEPARstart{W}{hen} sensing the world around us with an array of elements, we perform a spatio-temporal filtering operation known as beamforming. The conventional form of beamforming is delay-and-sum (DAS), where the signal from each element, $\vec{x}_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
The weight $\vec{w}$, is usually applied to trade resolution for a better signal-to-noise (SNR) ratio. Hence, we increase the main lobe width of the system while gaining a lower side lobe level.

The basic idea of the Capon beamformer is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria. For a base-band data vector $\vec{x}_m \in \mathbb{C}^M$, we now have a data-dependent weight, $\vec{w}(\vec{x}[n])$ that when applied on $\vec{x}$, is the best approximation to the selected criteria. Using the Capon criterion, the adaptive weights are found by solving the following minimization problem:
\begin{align}
&\min_{\vec{w}} E\{|z[n]|^2\} = \vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
&\text{subjected to } \vec{w}^H\vec{1} = 1,\label{eq:constraint}
\end{align}
where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the signal covariance matrix.
Thus, we minimize the output power, while maintaining unit gain in the look direction. The constraint in Eq. \ref{eq:constraint} is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons at broadside. The solution to the minimization problem in Eq. \ref{eq:minz} is
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
\end{align}
where $\vec{a} = \vec{1}$ when $\vec{x}$ are pre-delayed to broadside. In practice, the covariance matrix has to be approximated by the sample covariance matrix, $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$. In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over all available samples to improve on the estimate. For an active, pulsed system like ultrasound imaging, the available data from which $\mat{R}$ can be estimated is quite limited. The resulting sample covariance matrix might therefore be an inaccurate estimate, and in addition poorly conditioned. Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favors weights with opposite phase. In this way, the beamformer output can be zero at broadside (with signal pressent), even with the distortion-less criterion applied.

In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ subarrays and $N$ time samples.
\begin{align}
\mat{\hat{R}} = \frac{1}{NK}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,
\end{align}
where $K = M-L+1$, $N = 2N_{avg} + 1\sim \tau/T_s$ ($\tau$ is the pulse length and $T_s$ is the sampling period) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, $\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}$. A weighting proportional to the output power is beneficial, and the trace of $\mat{\hat{R}}$, $\epsilon = d*tr\{\mat{\hat{R}}\}$, has previously shown good results in the literature (cite).

Building and calculating the inverse of $\hat{R}$ is computational demanding, limiting the methods accessibility. Following the innovation in GPU computing, this has now started to change. 
In this paper we introduce the first GPU implementation of the Capon beamformer capable of processing a 70 degree sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing. Both types of smoothing are important for maintaining delay-and-sum-like speckle and at the same time keep the high resolution property. This is achieved using complex base band data, which reduces the size of each range line to a minimum. In addition, complex data results in complex weights; hence this is a correct implementation of Capon beamforming with all degrees of freedom preserved. For high channel count systems we propose two ways of reducing computational complexity while maintaining

\section{GPU Compute Model}
This turned out to be to technical?
In this section we briefly discuss the principle behind GPU computing. CUDA by Nvidia has been our choice of GPU developer environment. However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper will apply for other massive parallel implementation environments like e.g. OpenCL. The CUDA compute model is based on execution of a kernel across a grid of compute threads. Each position in the grid holds a block of threads. Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. From a CUDA perspective the GPU consist of one or more stream multi processors (SM), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions. The CUDA 2.x compute capability (CC) has available 48 KB of near-core shared memory per SM that can be utilized by resident blocks. The maximum number of resident blocks per SM is 8, but in addition the maximum number of resident warps per SM is restricted to 48. Hence, with a block size of 512 only $48/(512/32) = 3$ blocks can occupy the SM at once. In addition, if one thread where to use 64B of shared memory the number of allowed resident blocks with a block size of 512 would be reduced to 1. Up to a curtain point it is beneficial to have as many warps as possible occupying the SM at once to hide latency from accessing memory. The SM has for CC 2.x 32K 32-bit long registers. When implementing algorithms on the GPU it is important to keep data as close to the core as possible. Utilizing available registers and shared memory to its maximum is therefore important.

@@

\section{Parallel Capon}\label{sec:meth}
Capon beamforming has recently shown promising results in several studies. Although it improves the lateral resolution, widespread use has been limited by its computational complexity. However, the calculation of Capon weights is well suited for a GPU (Graphics Processing Unit) implementation, since the calculation is both independent across pixels, and partially intra-pixel.

\subsection{Building Multiple Sample Covariance Matrices}
Building $\hat{\mat{R}}$ using a sliding window approach across $K$ and $N_{avg}$. We have a limited amount of fast, near-core memory on the GPU. On NVIDIA architecture this memory is known as shared memory, and is restricted to 48KB per compute block. The maximum block size of 32x32, further restrict how large arrays we can handle inside a single compute block. It is therefore natural to restrict $L$ to a maxima of $32$ elements. Since $L \le M/2$, $M \le 64$. We can not afford to hold the full $\hat{\mat{R}}$ ($M = L$) when $\hat{\mat{R}} > 32$, because of limited amount of memory. ... Explain in detail how \texttt{buildR} is implemented in a kernel. ... Give more details on how much shared memory we have, and how it can be divided between compute blocks occupying one stream multiprocessor (SM).
